---
title: "Week2&3_solution"
author: "Xiao Cui"
date: "2024-04-02"
output: html_document
---

## Exercise A
### Task 1: Import data
```{r setup, include=FALSE} 
knitr::opts_chunk$set(warning = FALSE, message = FALSE) 
```

```{r library and load data}
# always work within a rproject
library(readr)
library(data.table)
library(dplyr)
library(ggplot2)
library(sf)
library(xfun)
library(tmap)
library(tidyr)
library(mapview)
wildschwein_BE_df <- read_csv("wildschwein_BE_2056.csv",
                 col_types = cols(
                   TierID = col_character(),
                   TierName = col_character(),
                   CollarID = col_integer(),
                   DatetimeUTC = col_datetime(format = "%Y-%m-%dT%H:%M:%SZ"),
                   Lat = col_double(),
                   Long = col_double()
                 ))

wildschwein_BE_df %>% as_tibble()
```
Now we have a look on the point patterns.
```{r plot points}
ggplot(wildschwein_BE_df, aes(E, N, colour = TierID)) + geom_point()
```


```{r convert coordinators to sf}
library("sf")
wildschwein_BE_sf <- st_as_sf(wildschwein_BE_df,
    coords = c("E", "N"),
    crs = 2056
)
plot(wildschwein_BE_sf)
```
### Task 2: Get an overview
```{r add time lag in df}
# define difftime function
difftime_secs <- function(later, now){
    as.numeric(difftime(later, now, units = "secs"))
}

# identify time lag within group (objects)
wildschwein_BE_sf <- group_by(wildschwein_BE_sf, TierID)
wildschwein_BE_sf <- mutate(wildschwein_BE_sf, timelag = difftime_secs(lead(DatetimeUTC), DatetimeUTC))

wildschwein_BE_sf
```
Q: How many individuals were tracked?
```{r task 2 question 1}
## individuals by type
name_ind <- unique(wildschwein_BE_sf$TierName)
numb_ind <- length(name_ind)

# Print (refer to codes from Tanja)
cat(crayon::bold("How many individuals were tracked?"), "\n")
cat("Number tracked individuals:", numb_ind, "\n")
cat("Tracked individuals:", name_ind, "\n")

```
A: There are 51246 individuals tracked.

Q2: For how long were the individual tracked? Are there gaps?
```{r task 2 question 2-1}
# Calculate total duration of tracking and gaps within each group
tracking_summary <- summarise(wildschwein_BE_sf, 
                              total_duration = difftime(max(DatetimeUTC), min(DatetimeUTC), units = "secs"))

tracking_summary
```

```{r task 2 question 2-2}
# Extract time component from DatetimeUTC
wildschwein_BE_sf$Time <- format(wildschwein_BE_sf$DatetimeUTC, format = "%H:%M:%S")
wildschwein_BE_sf$hour <- hour(wildschwein_BE_sf$DatetimeUTC)

# plot time distribution of individual tracking
ggplot(wildschwein_BE_sf, aes(x = as.Date(DatetimeUTC), y = hour, color = TierID)) +
  geom_point(alpha = 0.1) +
  labs(title = "Tracking Duration Over Time by TierID",
       x = "Date",
       y = "Time",
       color = "TierID") +
  theme_minimal()
```

A:
002A	29253602 secs			
016A	20275167 secs			
018A	22607072 secs

Q: Were all individuals tracked concurrently or sequentially?
```{r task 2 question 3}
# Check if individuals within each group were tracked with the count of unique timestamp
concurrent_tracking <- summarise(wildschwein_BE_sf, 
                                  tracked_concurrently = length(unique(DatetimeUTC)) == n(),
                                  n = n())
# show tracking time for each group
concurrent_tracking
```
A: Here we can find 016A and 018A have the similar total timestamp count. 002A has more timestamps than them, which means they were tracked sequentially. We can also refer to the time plot at the last part, i.e., there are obvious different patterns in time among each individual.

Q: What is the temporal sampling interval between the locations?
```{r task 2 question 4}
# Calculate temporal sampling interval within each group
sampling_interval <- summarise(wildschwein_BE_sf, 
                               sampling_interval = median(timelag, na.rm = TRUE))

# show temporal sampling
sampling_interval
```

A: The sampling interval here is around 900 second.

### Task 3: Get an overview
```{r task 3}
# Function to calculate step length between consecutive points
calculate_step_length <- function(geom) {
  later <- lag(geom)
  now <- geom
  
  # Function to calculate distance by time lag
  distance_by_element <- function(later, now) {
    as.numeric(
      st_distance(later, now, by_element = TRUE)
    )
  }
  
  # Calculate step length for each pair of consecutive points
  step_lengths <- distance_by_element(later, now)
  
  return(step_lengths)
}

# Example usage:
# Add step lengths as a new column to the sf object
wildschwein_BE_sf$steplength <- calculate_step_length(wildschwein_BE_sf$geometry)
wildschwein_BE_sf |> as_tibble()
```

The code above is an improvement from below:
```{r (not run) backup 1.1, eval = FALSE}
later <- lag(wildschwein_BE_sf$geometry)
now <- wildschwein_BE_sf$geometry

# calculate the distance by time lag
distance_by_element <- function(later, now){
  as.numeric(
    st_distance(later, now, by_element = TRUE)
  )
}

# Calculate step length for each pair of consecutive points
step_lengths <- distance_by_element(later, now)

# Add step lengths as a new column to the dataframe
wildschwein_BE_sf$steplength <- step_lengths
```

The initial version of this function:
```{r (not run) backup 1.0, eval = FALSE}
# code above is an improvement from:
wildschwein_BE_sf <- mutate(wildschwein_BE_sf, steplength = distance_by_element(later, now))
```
The reason is: Error: C stack usage  15922896 is too close to the limit.

### Task 4: Deriving distance and speed
```{r speed calculation}
# Calculate animals' speed based on steplength and timelag
# v = s/t
wildschwein_BE_sf$speed <- wildschwein_BE_sf$steplength / wildschwein_BE_sf$timelag

wildschwein_BE_sf |> as_tibble()
```


```{r sample and mapping}
# take a sample from "Sabi"
wildschwein_sample <- wildschwein_BE_sf |>
  filter(TierName == "Sabi") |> 
  head(100)

# create a map for our sample
tmap_mode("view")
tm_shape(wildschwein_sample) + 
  tm_dots()
```

```{r osm mapping}
# key parameter here: summarise(do_union = FALSE)
# connect points by order
wildschwein_sample_line <- wildschwein_sample |> 
  # dissolve to a MULTIPOINT:
  summarise(do_union = FALSE) |> 
  st_cast("LINESTRING")

tmap_options(basemaps = "OpenStreetMap")

tm_shape(wildschwein_sample_line) +
  tm_lines() +
  tm_shape(wildschwein_sample) + 
  tm_dots()
```

## Exercise B
```{r preparation}
# reload functions: calculate distance via time lag
difftime_secs <- function(x, y){
  as.numeric(difftime(x, y, units = "secs"))
}

distance_by_element <- function(later, now){
  as.numeric(
    st_distance(later, now, by_element = TRUE)
  )
}

calculate_movement_metrics <- function(df, datetime_col, n, timelag_col = "timelag", steplength_col = "steplength", speed_col = "speed") {
  now <- df[[datetime_col]]
  later <- now + 120*n # a sampling window of 120 seconds
  
  df[[timelag_col]] <- difftime(later, now, units = "secs")
  
  now_geo <- df$geometry
  later_geo <- lag(now_geo, n = n*2)
  
  df[[steplength_col]] <- distance_by_element(later_geo, now_geo)
  
  # Convert timelag to numeric
  df[[timelag_col]] <- as.numeric(df[[timelag_col]])
  
  df[[speed_col]] <- df[[steplength_col]] / df[[timelag_col]]
  
  return(df)
}
```
Then we import cow data.
```{r load cow data}
# import data
cow_df <- read_csv("caro60.csv",
                 col_types = cols(
                   TierID = col_character(),
                   TierName = col_character(),
                   CollarID = col_integer(),
                   DatetimeUTC = col_datetime(format = "%Y-%m-%dT%H:%M:%SZ"),
                   E = col_double(),
                   N = col_double()
                 ))
# convert dataframe to sf
cow_sf <- st_as_sf(cow_df, coords = c("E", "N"), crs = 2056)
# show sf breifly
cow_sf %>% as_tibble()
# keep time for calculation
caro <- cow_sf |> 
  select(DatetimeUTC)
# show caro briefly
caro %>% as_tibble()
```
### Task 1: Calculate speed at scale 1
```{r}
caro <- calculate_movement_metrics(caro, datetime_col = "DatetimeUTC", n = 1, timelag_col = "timelag1", steplength_col = "steplength1", speed_col = "speed1")

caro |> as_tibble()
```
### Task 2: Calculate speed at scale 2
```{r}
caro <- calculate_movement_metrics(caro, datetime_col = "DatetimeUTC", n = 2, timelag_col = "timelag2", steplength_col = "steplength2", speed_col = "speed2")

caro |> 
  # drop geometry and select only specific columns
  # to display relevant data only
  st_drop_geometry() |> 
  select(timelag2, steplength2, speed2) |> 
  head()
```

### Task 3: Calculate speed at scale 3
```{r}
caro <- calculate_movement_metrics(caro, datetime_col = "DatetimeUTC", n = 4, timelag_col = "timelag3", steplength_col = "steplength3", speed_col = "speed3")

caro |> 
  st_drop_geometry() |> 
  select(timelag3, steplength3, speed3) |> 
  head()
```
### Task 4: Compare speed across scales
```{r}
caro |> 
  st_drop_geometry() |> 
  select(DatetimeUTC, speed1, speed2, speed3)
```
### Task 5:
```{r plot speed 1}
ggplot(caro, aes(y = speed1)) + 
    # we remove outliers to increase legibility, analogue
  # Laube and Purves (2011)
  geom_boxplot(outliers = FALSE)
```
```{r compare speeds table}
# before pivoting, let's simplify our data.frame
caro2 <- caro |> 
  st_drop_geometry() |> 
  select(DatetimeUTC, speed1, speed2, speed3)

caro_long <- caro2 |> 
  pivot_longer(c(speed1, speed2, speed3))
  
head(caro_long)

ggplot(caro_long, aes(name, value)) +
  # we remove outliers to increase legibility, analogue
  # Laube and Purves (2011)
  geom_boxplot(outliers = FALSE)
```


## Exercise C
We export our trajectory from posmo and import data here.
```{r import posmo data}
posmo_df <- read_csv("posmo_03_23_05_03.csv",
                 col_types = cols(
                   user_id = col_character(),
                   datetime = col_datetime(),
                   weekday = col_factor(levels = c("Sun", "Mon", "Tue", "Wed", "Thu", "Fri", "Sat")), # read daytime as a factor
                   place_name = col_character(),
                   transport_mode = col_character(),
                   lon_x = col_double(),
                   lat_y = col_double()
                 ))

posmo_df %>% as_tibble()
```
```{r posmo sf}
# Convert to sf
posmo_sf <- st_as_sf(posmo_df, coords = c("lon_x", "lat_y"), crs = 4326)
# Convert CRS to CH1903 / LV95 (EPSG:2056)
posmo_sf <- st_transform(posmo_sf, crs = 2056)
```
There are too many point data in posmo_sf. Thus, we will use sample to have a look at our data. We take trajectory on April 21 as a sample here.
Wrap up: we need the package data.table. It converts S3: POSIXct into string. I am flexible if there is a potential better method for filtering date. Examples are shown below, also see: https://cran.r-project.org/web/packages/data.table/vignettes/datatable-intro.html
https://stackoverflow.com/questions/9839343/extracting-time-from-posixct
```{r (not run) backup time format, eval = FALSE}
library(data.table)
x <- "2012-03-07 03:06:49 CET"
as.IDate(x) # Output is "2012-03-07"
hour(x): extract hour from timestamp
```

```{r mapping dots by time}
# take a sample from "2024-04-03"
posmo_0304_sf <- posmo_sf |>
  filter(as.IDate(datetime) == "2024-04-03")

# a function for assigning colors
assign_color <- function(x) {
  # Extract hour from datetime column
  hour <- data.table::hour(x$datetime)
  
  # Assign colors based on hour ranges
  color <- case_when(
    hour >= 6 & hour < 12 ~ "orange",
    hour >= 12 & hour < 18 ~ "blue",
    TRUE ~ NA_character_
  )
  
  # Create a summary data frame with colors
  color_summary <- data.frame(geometry = x, color = color)
  
  # Convert to sf object
  color_sf <- st_sf(color_summary, crs = st_crs(x))
  
  return(color_sf)
}

# Create colored sf object
posmo_0304_sf_color <- assign_color(posmo_0304_sf)

# repeated steps in Exercise A
# Create the map with colored point data by time
tmap_mode("view")
tm_shape(posmo_0304_sf_color) + 
  tm_dots(col = "color")

# key parameter here: summarise(do_union = FALSE)
# connect points by order
posmo_0304_line <- posmo_0304_sf |> 
  # dissolve to a MULTIPOINT:
  summarise(do_union = FALSE) |> 
  st_cast("LINESTRING")

# use tmap for visualizing our trajectory
tmap_options(basemaps = "OpenStreetMap")

tm_shape(posmo_0304_line) +
  tm_lines() +
  tm_shape(posmo_0304_sf) + 
  tm_dots()
```






